---
title: "Exploring gapminder: Solution to final challenge"
author: "Naupaka Zimmerman"
date: "April 19, 2015"
output: html_document
---

```{r load-libraries-and-data, echo=FALSE, message=FALSE, warning=FALSE}
# Load dplyr for data manipulation
library("dplyr")

# Load knitr so we can use the kable() function to format the table output more 
# nicely in our report below
library("knitr")

# Read in the gapminder tab-delimited data
data.in <- read.table("output/combined_gapMinder.tsv", 
                      sep = "\t", header = TRUE)
```

```{r define-scatterplot-function, echo=FALSE}
# function to make a scatter plot of year by GDP per capita for a given country
country.scatter.plot <- function(country.name){
    # first filter the data to just the desired rows
    data.in %>%
        filter(country == country.name) -> filtered.df

    # then plot those rows
    plot(x = filtered.df$year, 
         y = filtered.df$gdpPercap,
         main = paste("GDP by year in", country.name),
         xlab = "Year",
         ylab = "GDP per capita")
}
```

### Scatter plots 

```{r plot-china-scatter, echo=FALSE}
country.scatter.plot("China")
```

We see a huge increase in per capita GDP in China starting in the 1990s.

```{r plot-cuba-scatter, echo=FALSE}
country.scatter.plot("Cuba")
```

In Cuba, GDP reaches a peak in the 1980s, before becoming much lower in the 1990s, and climbing again after the year 2000.

```{r plot-chile-scatter, echo=FALSE}
country.scatter.plot("Chile")
```

The GDP per capita in Chile increases continually, especially after 1990, like China, but in absolute terms, GDP  in Chile is more than twice as high than in China over the same time period.

### Summary statistics

```{r calc-mean-min-max-continents, echo = FALSE}
# Calculate the mean, min, and max for each continent and store to a data frame
data.in %>%
    group_by(continent) %>%
    summarize(mean = mean(lifeExp), min = min(lifeExp), max = max(lifeExp)) -> summarized.df
```

```{r display-summary-table, results='asis', echo = FALSE}
# the kable() function turns a data.frame object into a nicely formatted markdown table
# that we can include in our html report. We need the as-is chunk option because the output 
# of the function is markdown and we want this to go directly into the report
kable(summarized.df, caption = "Life expectancies per continent")
```

```{r function-for-in-text-values, echo = FALSE}
summary.value <- function(continent, column){
    # this function takes a continent name and a column name (e.g. mean, max, or min)
    # and returns that number formatted to only include 4 digits (using the format() function) 
    # since we don't want or need the precision here for reporting
    format(summarized.df[summarized.df$continent == continent, column], digits = 4)
}
```

Mean life expectancies (MLE) over this period are highest in Oceania, where MLE is `r summary.value("Oceania", "mean")`. They are lowest in Africa (mean `r summary.value("Africa", "mean")`). While maximum life expectancies are relatively similar across continents, minimum life expectancies vary dramtically, and the lowest values are from Africa (`r summary.value("Africa","min")`) and Asia (`r summary.value("Asia","min")`).

### Histograms of global life expectancy

```{r plot-default-bins-hist, echo=FALSE}
# defaults to 14 breaks in this case
hist(data.in$lifeExp, 
     main = "Distribution of global life expectancy",
     xlab = "Life expectancy")
```

We see a peak in life expectancy around 70 years of age, although it also seems that the distribution is non-normal.

```{r plot-addtl-bins-hist, echo=FALSE}
# We'll increase the number of breaks to 100 to get finer resolution
hist(data.in$lifeExp, breaks = 100, 
     main = "Distribution of global life expectancy",
     xlab = "Life Expectancy")
```

The non-normality is even more evident with more bins, and there is even a hint that it might be bi-modal, with a second (less frequent) peak near 45 years of age.
